Model Comparisons

### 2015-11-26 models/Model_Comparison.Rmd
model comparisons with normalization adjustments and subjective alts from e10a.csv

# Run model functionality
```{r}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
#     mutate(hi = hi / sum(hi), 
#            low = low / sum(low)) %>%
    select(hi, low)

@@

# run.full()
# ------------
# Run RSA with model3 (full) with alternatives
run.full = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
#     mutate(hi1 = hi1 / sum(hi1),
#            hi2 = hi2 / sum(hi2),
#            low1 = low1 / sum(low1),
#            low2 = low2 / sum(low2)) %>%
    select(hi1, hi2, low1, low2)

@@

## Tune hyperparams
```{r}
# tune.alhpa()
# ------------
# d        --> data
# alphas   --> range of alphas to test
# type     --> full or partial model
# useprior --> use uniform prior
# usenone  --> use gener None
tune.alpha = function(d, alphas = seq(from=1, to=10), type="partial", useprior = T, usenone=F, compare.data=NULL) {
  # Tune best alphas
  fit = sapply(alphas, FUN=function(n) {
    if (type == "partial") {
      md = d %>%
        do(run.partial(., alpha = n, useprior = useprior, usenone = usenone))
    } else {
      md = d %>%
        do(run.full(., alpha = n, useprior = F, usenone = F))
    }
    # Fit to e6 data
    if (!is.null(compare.data)) {
      matched.items = which((md[, "scale"] != "some_all" &
                               (md[, "degree"] == "hi2" | md[, "degree"] == "hi1")) |
                              (md[, "scale"] == "some_all" &
                                 (md[, "degree"] == "hi1" | md[, "degree"] == "low1")))
      md = md[matched.items, ]
      md$degree = ifelse(md$degree == "hi1", "hi", "low")
      stopifnot("listener.p" %in% colnames(compare.data))
      md$listener.p = compare.data$listener.p
    }

@@

## Data for full model
```{r full_model_data, message=FALSE, warning=FALSE}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10.csv")
listener = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L1_e11.csv")
data.full = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

## Match items between studies (full and partial)
```{r}
matched.items = which((data.full[, "scale"] != "some_all" &
        (data.full[, "degree"] == "hi2" | data.full[, "degree"] == "hi1")) |
       (data.full[, "scale"] == "some_all" &
          (data.full[, "degree"] == "hi1" | data.full[, "degree"] == "low1")))
```

## Run comparisons
```{r run_comparisons, message=FALSE}
# Save performance output
performance.output = data.frame(model=rep(NA, 6), cor=rep(NA, 6))
alphas = rep(NA, 6)

@@

# Model 3 - Full, no alpha tuning
# -------------------------------
alphas[5] = 1
m3.noFit = data.full %>%
  do(run.full(., alpha = alphas[5]))
m3.noFit.matched = m3.noFit[matched.items, ]
m3.noFit.matched$degree = ifelse(m3.noFit.matched$degree == "hi1", "hi", "low")
m3.noFit.matched = cbind(m3.noFit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.noFit.matched$scale == data.partial$scale & m3.noFit.matched$degree == data.partial$degree)
colnames(m3.noFit.matched)[length(colnames(m3.noFit.matched))] = "e6.listener.p"
performance.output[5, ] = c("M3_noFit", round(cor(m3.noFit.matched$e6.listener.p, m3.noFit.matched$pred), 5))

# Model 3 - Full, alpha tuning
# ----------------------------
alphas[6] = tune.alpha(data.full, type = "full", compare.data=data.partial)
m3.fit = data.full %>%
  do(run.full(., alpha = alphas[6]))
m3.fit.matched = m3.fit[matched.items, ]
m3.fit.matched$degree = ifelse(m3.fit.matched$degree == "hi1", "hi", "low")
m3.fit.matched = cbind(m3.fit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.fit.matched$scale == data.partial$scale & m3.fit.matched$degree == data.partial$degree)
colnames(m3.fit.matched)[length(colnames(m3.fit.matched))] = "e6.listener.p"
performance.output[6, ] = c("M3_fit", round(cor(m3.fit.matched$e6.listener.p, m3.fit.matched$pred), 5))
```

@@

Overall model performance: $r^2$. M1 and M2 use literal listener values from exp8. M2 includes a generic "None" defined in terms of stars (1: 1.0, 2:0.0, 3:0.0, 4:0.0, 5:0.0). M3 uses the full set of literal listener alternatives from exp10, however r^2 values and tuning reflects comparisons to exp6 pragmatic listener judgments (not exp11).
```{r overall_perf}
performance.output$model = factor(performance.output$model, levels=c(
  "M1_noFit", "M1_fit", "M2_noFit", "M2_fit", "M3_noFit", "M3_fit"
))
# Populate extra model info
performance.output$cor = as.numeric(performance.output$cor)
performance.output$alphas = alphas
performance.output$alts = c(rep("entailment only", 2), rep("entailment + None", 2), rep("full", 2))
grid.table(performance.output)

@@

m3.matched.noFit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m3.noFit.matched,
      main=paste("M3, no fit\nalpha: ", alphas[5]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.noFit.G

m3.matched.fit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m3.fit.matched,
      main=paste("M3, fit\nalpha: ", alphas[6]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.fit.G

@@

# Full model fitted - color stars
m3.matched.fit.CorPlot = ggplot(data=m3.fit.matched, aes(y=pred, x=e6.listener.p)) +
  geom_point(aes(colour = stars)) +
  geom_smooth(method=lm) +
  ggtitle("Model performance: M3 fit") +
  labs(x = "human judgments", y="model prediction")
m3.matched.fit.CorPlot
```

>>>>

# Run model functionality
```{r}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial = function(d, alpha = 1, useprior = F, usenone = F, normalize=F) {
  if (normalize) {
    mat = d %>%
      select(stars, degree, speaker.p) %>%
      spread(degree, speaker.p) %>%
      mutate(hi = hi / sum(hi),
             low = low / sum(low)) %>%
      select(hi, low)
  } else {
    mat = d %>%
      select(stars, degree, speaker.p) %>%
      spread(degree, speaker.p) %>%
      select(hi, low)
  }

@@

# run.full()
# ------------
# Run RSA with model3 (full) with alternatives
run.full = function(d, alpha=1, useprior=F, usenone=F, addMid=F, normalize=F) {
  if (addMid) {
    if (normalize) {
      mat = d %>%
        select(stars, degree, speaker.p) %>%
        spread(degree, speaker.p) %>%
        mutate(hi1 = hi1 / sum(hi1),
               hi2 = hi2 / sum(hi2),
               mid = mid / sum(mid),
               low1 = low1 / sum(low1),
               low2 = low2 / sum(low2)) %>%
        select(hi1, hi2, mid, low1, low2)    
    } else {
      mat = d %>%
        select(stars, degree, speaker.p) %>%
        spread(degree, speaker.p) %>%
        select(hi1, hi2, mid, low1, low2)    
    }
  } else {
    if (normalize) {
      mat = d %>%
        select(stars, degree, speaker.p) %>%
        spread(degree, speaker.p) %>%
        mutate(hi1 = hi1 / sum(hi1),
               hi2 = hi2 / sum(hi2),
               low1 = low1 / sum(low1),
               low2 = low2 / sum(low2)) %>%
        select(hi1, hi2, low1, low2)
    } else {
      mat = d %>%
        select(stars, degree, speaker.p) %>%
        spread(degree, speaker.p) %>%
        select(hi1, hi2, low1, low2)
    }
  }

@@

## Tune hyperparams
```{r}
# tune.alhpa()
# ------------
# d        --> data
# alphas   --> range of alphas to test
# type     --> full or partial model
# useprior --> use uniform prior
# usenone  --> use gener None
tune.alpha = function(d, alphas = seq(from=1, to=10),
                      type="partial", useprior = T,
                      usenone=F, compare.data=NULL, addMid = F, normalize=F) {
  # Tune best alphas
  fit = sapply(alphas, FUN=function(n) {
    if (type == "partial") {
      md = d %>%
        do(run.partial(., alpha = n, useprior = useprior, usenone = usenone, normalize=normalize))
    } else {
      md = d %>%
        do(run.full(., alpha = n, useprior = useprior, usenone = usenone, addMid=addMid, normalize=normalize))
    }
    # Fit to e6 data
    if (!is.null(compare.data)) {
      matched.items = which((md[, "scale"] != "some_all" &
                               (md[, "degree"] == "hi2" | md[, "degree"] == "hi1")) |
                              (md[, "scale"] == "some_all" &
                                 (md[, "degree"] == "hi1" | md[, "degree"] == "low1")))
      md = md[matched.items, ]
      md$degree = ifelse(md$degree == "hi1", "hi", "low")
      stopifnot("listener.p" %in% colnames(compare.data))
      md$listener.p = compare.data$listener.p
    }

@@

## Data for full model
```{r full_model_data, message=FALSE, warning=FALSE}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10.csv")
listener = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L1_e11.csv")
data.full = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)

speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10a.csv")
data.full.extras = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

## Match items between studies (full and partial)
```{r}
matched.items = which((data.full[, "scale"] != "some_all" &
        (data.full[, "degree"] == "hi2" | data.full[, "degree"] == "hi1")) |
       (data.full[, "scale"] == "some_all" &
          (data.full[, "degree"] == "hi1" | data.full[, "degree"] == "low1")))
matched.items.extras = which((data.full.extras[, "scale"] != "some_all" &
        (data.full.extras[, "degree"] == "hi2" | data.full.extras[, "degree"] == "hi1")) |
       (data.full.extras[, "scale"] == "some_all" &
          (data.full.extras[, "degree"] == "hi1" | data.full.extras[, "degree"] == "low1")))
```

## Run comparisons
```{r run_comparisons, message=FALSE}
# Save performance output
performance.output = data.frame(model=rep(NA, 8), cor=rep(NA, 8))
alphas = rep(NA, 8)

@@

# Model 3 - Full, no alpha tuning
# -------------------------------
alphas[5] = 1
m3.noFit = data.full %>%
  do(run.full(., alpha = alphas[5], addMid=F, normalize=F))
m3.noFit.matched = m3.noFit[matched.items, ]
m3.noFit.matched$degree = ifelse(m3.noFit.matched$degree == "hi1", "hi", "low")
m3.noFit.matched = cbind(m3.noFit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.noFit.matched$scale == data.partial$scale & m3.noFit.matched$degree == data.partial$degree)
colnames(m3.noFit.matched)[length(colnames(m3.noFit.matched))] = "e6.listener.p"
performance.output[5, ] = c("M3_noFit", round(cor(m3.noFit.matched$e6.listener.p, m3.noFit.matched$pred), 5))

# Model 3 - Full, alpha tuning
# ----------------------------
alphas[6] = tune.alpha(data.full, type = "full", compare.data=data.partial, addMid=F, normalize=F)
m3.fit = data.full %>%
  do(run.full(., alpha = alphas[6]))
m3.fit.matched = m3.fit[matched.items, ]
m3.fit.matched$degree = ifelse(m3.fit.matched$degree == "hi1", "hi", "low")
m3.fit.matched = cbind(m3.fit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.fit.matched$scale == data.partial$scale & m3.fit.matched$degree == data.partial$degree)
colnames(m3.fit.matched)[length(colnames(m3.fit.matched))] = "e6.listener.p"
performance.output[6, ] = c("M3_fit", round(cor(m3.fit.matched$e6.listener.p, m3.fit.matched$pred), 5))

# Model 4 - Full, no alpha tuning
# -------------------------------
alphas[7] = 1
m4.noFit = data.full.extras %>%
  do(run.full(., alpha = alphas[7], addMid=T))
m4.noFit.matched = m4.noFit[matched.items.extras, ]
# Need to fix this to dealt with 'mid'
m4.noFit.matched$degree = ifelse(m4.noFit.matched$degree == "hi1", "hi", "low")
m4.noFit.matched = cbind(m4.noFit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m4.noFit.matched$scale == data.partial$scale & m4.noFit.matched$degree == data.partial$degree)
colnames(m4.noFit.matched)[length(colnames(m4.noFit.matched))] = "e6.listener.p"
# store corr
performance.output[7, ] = c("M4_noFit", round(cor(m4.noFit.matched$e6.listener.p, m4.noFit.matched$pred), 5))

# Model 4 - Full, alpha tuning, added 'mid'
# ----------------------------
alphas[8] = tune.alpha(data.full.extras, type = "full", compare.data=data.partial, addMid=T)
m4.fit = data.full.extras %>%
  do(run.full(., alpha = alphas[8], addMid=T))
m4.fit.matched = m4.fit[matched.items.extras, ]
m4.fit.matched$degree = ifelse(m4.fit.matched$degree == "hi1", "hi", "low")
m4.fit.matched = cbind(m4.fit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m4.fit.matched$scale == data.partial$scale & m4.fit.matched$degree == data.partial$degree)
colnames(m4.fit.matched)[length(colnames(m4.fit.matched))] = "e6.listener.p"
# store corr
performance.output[8, ] = c("M4_fit", round(cor(m4.fit.matched$e6.listener.p, m4.fit.matched$pred), 5))
```

@@

Overall model performance: $r^2$. M1 and M2 use literal listener values from exp8. M2 includes a generic "None" defined in terms of stars (1: 1.0, 2:0.0, 3:0.0, 4:0.0, 5:0.0). M3 uses the full set of literal listener alternatives from exp10, however r^2 values and tuning reflects comparisons to exp6 pragmatic listener judgments (not exp11).
```{r overall_perf}
performance.output$model = factor(performance.output$model, levels=c(
  "M1_noFit", "M1_fit", "M2_noFit", "M2_fit", "M3_noFit", "M3_fit", "M4_noFit", "M4_fit"
))
# Populate extra model info
performance.output$cor = as.numeric(performance.output$cor)
performance.output$alphas = alphas
performance.output$alts = c(rep("entailment only", 2), rep("entailment + None", 2), rep("full", 2), rep("full + extras", 2))
grid.table(performance.output)

@@

m3.matched.noFit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m3.noFit.matched,
      main=paste("M3, no fit\nalpha: ", alphas[5]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.noFit.G
cor(m3.noFit.matched$e6.listener.p, m3.noFit.matched$pred)

m3.matched.fit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m3.fit.matched,
      main=paste("M3, fit\nalpha: ", alphas[6]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.fit.G
cor(m3.fit.matched$e6.listener.p, m3.fit.matched$pred)

@@

# Full model fitted - color stars
m3.matched.fit.CorPlot = ggplot(data=m3.fit.matched, aes(y=pred, x=e6.listener.p)) +
  geom_point(aes(colour = stars)) +
  geom_smooth(method=lm) +
  ggtitle("Model performance: M3 fit") +
  labs(x = "human judgments", y="model prediction")
m3.matched.fit.CorPlot

m4.matched.noFit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m4.noFit.matched,
      main=paste("M4, no fit\nalpha: ", alphas[7]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m4.matched.noFit.G

m4.matched.fit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m4.fit.matched,
      main=paste("M4, fit\nalpha: ", alphas[8]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m4.matched.fit.G
```
